---
title: Accessing CEFI Cloud data with R
reticulate:
  python: "/srv/conda/envs/notebook/bin/python"
---
{{< downloadthis r_get_cefi_cloud_rstudio.qmd dname='r_get_cefi_cloud_rstudio.qmd' >}}

The following notebook is a quick demostration on how to use R to access the CEFI Cloud data. 

We will explore the process of extracting a subset of model simualtion data produce from the regional MOM6 model. The model output is to support the Climate Ecosystem and Fishery Initiative. We will showcase how to utilize xarray and fsspec for accessing the data and visualize it on an detailed map. The currently available data can be viewed on 

- [AWS](https://noaa-oar-cefi-regional-mom6-pds.s3.amazonaws.com/index.html)
- [Google Cloud Storage](https://console.cloud.google.com/storage/browser/noaa-oar-cefi-regional-mom6)

The contents of this folder encompass hindcast simulations derived from the regional MOM6 model spanning the years 1993 to 2019.

## Packages used

In this example, we are going to use the _reticulate_ library with Python to read the CEFI data into an R data structure. From there you can operate on the data in R (or R Studio) as you would any other data.

The other libraries are used to filter the data and make a plot on a detailed base map.

```{r message=FALSE}
library(reticulate)
library(dplyr)
library(maps)
library(ggplot2)
# reticulate::py_config()  # check python version
```

## Setting up the Python packages

There are 2 approaches and which one you use will depend on the environment you are using. 

First make sure you have `reticulate` R package installed.
```{r}
print(requireNamespace("reticulate", quietly = TRUE))
```
If needed, install.
```{r eval=FALSE}
install.packages("reticulate")
```

### You installed R with conda via `r-irkernel`

*This is the approach to use if you are using the "CEFI" environment.*

If you are using an environment where R is installed with conda via the `r-irkernel` package, you can do the following. 

```{r eval=FALSE}
# reticulate has a hard time finding conda
options(reticulate.conda_binary = "/srv/conda/condabin/conda")
reticulate::use_condaenv("/srv/conda/envs/notebook", required = TRUE)
```

Note, you might run into an issue where reticulate refuses to change the Python version. This can happen if you used reticulate without conda and it stored that Python path in the cache. 1) Session > Terminate R 2) Open terminal and run `rm -rf .cache/R/reticulate`.

### You installed R outside of conda

*This is the approach to use if you are using the "py-rocket-geospatial" environment (default on the NMFS JupyterHub).*

If you installed R like you normally would (not with conda), then you are liable to run into SSL issues when `reticulate` tries to open files in the cloud since `reticulate` will use a version of SSL installed with RStudio and not the one in the conda environment.

**Option 1. Do not use conda**

You do not need to use conda with reticulate. Use `py_require()` to create an ephemeral environment with your packages. Note there is a problem with `zarr 3.*` and opening zarr files, so you need to make sure `zarr<3` is installed.

```{r}
py_require(c("zarr<3", "fsspec", "xarray", "gcsfs", "s3fs"))
xr <- import('xarray')
fsspec <- import('fsspec')
```

**Option 3. Force RStudio to use the right SSL**

Let's say you want to use your conda environment but you are getting the SSL error. You need to force RStudio to use your conda SSL when RStudio starts up. You do this by adding a line to `/etc/rstudio/rserver.conf`. This only works if you have permission to write to `/etc/rstudio` or you can use `sudo`.

Before starting RStudio, open a terminal and run this. If you are in a Jupyter Hub, you will need to run this every time you start your server (your hub).
```
echo "rsession-ld-library-path=/srv/conda/envs/notebook/lib" >> /etc/rstudio/rserver.conf
```
Replace "notebook" with the name of your conda environment if it is different.

## Get some data from the cloud

`fsspec` and `xarray` are the Python packages we need to get data from the cloud.
We're going to import them into our R environment and use them to create a pandas dataframe with lat, lon, tos as the columns

- Read the data
- select a time slice


```{r}
fs <- fsspec$filesystem(
  "reference",
  remote_protocol = "gcs",
  fo = "gcs://noaa-oar-cefi-regional-mom6/northwest_atlantic/full_domain/hindcast/monthly/regrid/r20230520/tos.nwa.full.hcast.monthly.regrid.r20230520.199301-201912.json",
  remote_options = dict(anon = TRUE),
  skip_instance_cache = TRUE,
  target_options = dict(anon = TRUE)
)
m <- fs$get_mapper()
dataset <- xr$open_dataset(m, engine='zarr', consolidated=FALSE)
subset <- dataset$sel(time='2012-12')
```

Get the variable long_name, date of the slice and data set title for the plot below

```{r}
datetime <- subset$time$values
title <- subset$attrs$title
long_name <- subset$tos$attrs$long_name
```

## Reorganize that data into a data.frame 
We transform the gridded data into column data with lon, lat, tos as columns.
**Take care:** We are expanding along lon first and then by lat, so we reorder the xarray so that lon varies fastest to match.

```{r}
subset <- subset$transpose('lon','lat','time')
df <- expand.grid(x = subset$lon$values, y=subset$lat$values)
data <- as.vector(t(matrix(subset$tos$values)))
df$Data <- data
names(df) <- c("lon", "lat", "tos")
```

## Subset the dataset
We use filter to extract the points over the Gulf region.

```{r}
df_gulf = filter(df, lat>17)
df_gulf = filter(df_gulf, lat<31)
df_gulf = filter(df_gulf, lon>260)
df_gulf = filter(df_gulf, lon<290)
df_gulf
```

## Longitude Manipulation 

The model lons are from 0 to 360, so we're going to use this function to normalize them to -180 to 180
The map polygons are defined on -180 to 180 we need to adjust our values to see them on the map

```{r}
normalize_lon <- function(x) {
  quotient <- round(x / 360.)
  remainder <- x - (360. * quotient)
  # Adjust sign if necessary to match IEEE behavior
  if (sign(x) != sign(360.) && remainder != 0) {
    remainder <- remainder + sign(360.) * 360.
  }
  return(remainder)
}
df_gulf$lon <- unlist(lapply(df_gulf$lon, normalize_lon))
```

## Plot setup
Set up a resonable plot size and limits for the plot area of the world base map

```{r}
options(repr.plot.width = 10, repr.plot.height = 10)
ylim <- c( min(df_gulf$lat), max(df_gulf$lat) )
xlim <- c( min(df_gulf$lon), max(df_gulf$lon) )
```

## Mesh/Dot Map 
Plot a mesh/dot colored according to the value of tos at each lat/lon location in the Gulf.

```{r}
w <- map_data( 'world', ylim = ylim, xlim = xlim )
p = ggplot() + 
    geom_polygon(data = w, aes(x=long, y = lat, group = group), fill = "grey", color = "black") +
    coord_fixed(1.3, xlim = xlim, ylim = ylim) + 
    geom_point(data = df_gulf, aes(color = tos, y=lat, x=lon), size=.5) +
    labs(title = paste(long_name, 'at', datetime, 'of', title))
p
```

## Accessing the data on the original model grid
The original model grid is [explained in detail](rmom6Grid) at the beginning of the cookbook. Understanding where the data are located on the earth requires you to merge the raw grid data file with the grid description.
We'll get both the data and the grid from the cloud storage just as we did in the example above and merge them using xarray, then we'll convert them to a data.frame for plotting (or processing) in R.

**N.B.** we use the AWS cloud storage here. We used the Google Cloud Storage above. You can use either, but if you are computing in a cloud environment you'll get better performance if you use the same cloud.

**N.B.** The "bucket" names are slightly different

**N.B.** Additional parameters (remote_options and target_options) are needed to let AWS know that the access does not require any credentials and could be done anonymously.

```{r}
params = list(anon = TRUE)
fs <- fsspec$filesystem('reference', 
    remote_protocol='s3',
    fo='s3://noaa-oar-cefi-regional-mom6-pds/northwest_atlantic/full_domain/hindcast/monthly/raw/r20230520/sos.nwa.full.hcast.monthly.raw.r20230520.199301-201912.json',
    remote_options=params,
    target_options=params
)
m <- fs$get_mapper()
dataset <- xr$open_dataset(m, engine='zarr', consolidated=FALSE)
fsg <- fsspec$filesystem('reference', 
    remote_protocol='s3',
    fo='s3://noaa-oar-cefi-regional-mom6-pds/northwest_atlantic/full_domain/hindcast/monthly/raw/r20230520/ocean_static.json',
    remote_options=params,
    target_options=params
)
gm <- fsg$get_mapper()
dataset <- xr$open_dataset(m, engine='zarr', consolidated=FALSE)
grid <- xr$open_dataset(gm, engine='zarr', consolidated=FALSE)
```

### Slicing the data

Here, we get one month slice from the data set. We took June 2003 in this case.

```{r}
slice = dataset$sel(time='2003-06')
title = slice$attrs$title
long_name = slice$sos$long_name
datetime = slice$time$values
slice
```

```{r}
lon <- grid$geolon$values
lat <- grid$geolat$values
```

We're using the lat and lon values from the static grid file to pair with each data value we read from the raw grid file to create a data.frame with lon,lat,sos

```{r}
slice <- slice$transpose('xh', 'yh', 'time', 'nv')
X <- as.vector(t(lon))
Y <- as.vector(t(lat))
data <- as.vector(t(matrix(slice$sos$values)))

df <- data.frame(
  "lon" = X,
  "lat" = Y,
  "sos" = data
)
```

```{r}
sub = filter(df, lat>35)
sub = filter(sub, lat<45)
sub = filter(sub, lon>=-80)
sub = filter(sub, lon<=-60)
sub
```

```{r}
options(repr.plot.width = 10, repr.plot.height = 10)
ylim <- c( min(sub$lat), max(sub$lat) )
xlim <- c( min(sub$lon), max(sub$lon) )
```

```{r}
w <- map_data( 'world', ylim = ylim, xlim = xlim )
p = ggplot() + 
    geom_polygon(data = w, aes(x=long, y = lat, group = group), fill = "grey", color = "black") +
    coord_fixed(1.3, xlim = xlim, ylim = ylim) + 
    geom_point(data = sub, aes(color = sos, y=lat, x=lon), size=1) +
    labs(title = paste(long_name, 'at', datetime, 'of', title, '\nRaw Grid'))
p
```


